Action Groupoids and Fractional Types

### 2016-07-02 pibackground.tex

\noindent Both classes of programs, 1-combinators $c$ and
2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT
sense. The elements $c$ of 1-combinators denote type isomorphisms. The
elements $\alpha$ of 2-combinators denote equivalences between these
type isomorphisms. Using the 1-combinators, it is possible to write
any reversible boolean function and hence encode arbitrary boolean
functions by a technique that goes back to \citet{Toffoli:1980}. The
2-combinators provide a layer of programs that computes
semantic-preserving transformations of 1-combinators. As a small
example, let us abbreviate $\ot \oplus \ot$ as $\mathsf{Bool}$ and
examine two possible implementations of boolean negation. The first
directly uses the primitive combinator
$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to swap the
two values of type $\mathsf{Bool}$; the second use three consecutive
copies of $\swapp$ to achieve the same effect:
\[\begin{array}{rcl}
\mathsf{not_1} &=& \swapp \\
\mathsf{not_2} &=& (\swapp \odot \swapp) \odot \swapp 
\end{array}\]
We can write a 2-combinator whose \emph{type} is $\mathsf{not_2}
\isotwo \mathsf{not_1}$:
\[
(\idldl ~\respstwo~ \idisotwo)~\transtwo~\idldl
\]
which not only shows the equivalence of the two implementations of
negation but also shows \emph{how} to transform one to the other.

Fig.~\ref{pi-combinators} lists all the 1-combinators which consist of
base combinators (on the left) and compositions (on the right). Each
line of the base combinators introduces a pair of dual
constants\footnote{where $\swapp$ and $\swapt$ are self-dual.} that
witness the type isomorphism in the middle. This set of isomorphisms
is known to be sound and
complete~\cite{Fiore:2004,fiore-remarks}. Fig.~\ref{pi-combinators2}
lists a few of the 2-combinators that we use in this paper. Each
2-combinator relates two 1-combinators of the same type and witnesses
their equivalence. The 2-equivalences behave as expected with respect
to inverses of 1-combinators as shown below.

@@

We give an operational semantics for the 1-combinators of $\Pi$ which
represent ``conventional'' programs.  Operationally, the semantics
consists of a pair of mutually recursive evaluators that take a
combinator and a value and propagate the value in the forward
direction~$\triangleright$ or in the backward
direction~$\triangleleft$. We show the complete forward evaluator in
Fig.~\ref{opsem}; the backward evaluator is easy to infer.

@@

\paragraph{Intermezzo.} For an information-theoretic perspective on
the language $\Pi$, we think of a type containing $N$ values as an
abstract system that has~$N$ distinguishable states. According to the
conventional theory of information~\cite{Shannon1948}, the amount of
information contained in each state of a system with $N$
distinguishable states is $\log N$ bits of information. For example,
the type $\mathsf{Bool}$ can be thought of as an abstract system with
two distinguishable states labeled $\mathsf{true}$ and
$\mathsf{false}$ each containing $\log 2 = 1$ bit. Similarly, the type
$\mathbb{3}$ can be thought of as an abstract system with three
distinguishable states each containing $\log 3$ bits. The logarithmic
map implies that information contained in a composite state is the sum
of the information contained in its constituents. For example, the
type $\mathsf{Bool} \otimes \mathbb{3}$ can be thought of a composite
system consisting of two independent unrelated subsystems. Each state
of the composite system therefore contains
$\log (2 * 3) = \log 2 + \log 3$ bits which is the sum of the
information contained in each subsystem. Since all the combinators
preserve cardinality, they are also information-preserving.

>>>>

\noindent Both classes of programs, 1-combinators $c$, and
2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT
sense. The elements $c$ of 1-combinators denote type isomorphisms. The
elements $\alpha$ of 2-combinators denote equivalences between these
type isomorphisms which are sound and complete 1-combinator program
transformations. Using the 1-combinators, it is possible to write any
reversible boolean function and hence encode arbitrary boolean
functions by a technique that goes back to \citet{Toffoli:1980}. The
2-combinators provide a layer of programs that computes
semantic-preserving transformations of 1-combinators. As a small
example, let us abbreviate $\ot \oplus \ot$ as the type $\mathbb{2}$
of booleans and examine two possible implementations of boolean
negation. The first directly uses the primitive combinator
$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to swap the
two values of type $\mathbb{2}$; the second use three consecutive
copies of $\swapp$ to achieve the same effect:
\[\begin{array}{rcl}
\mathsf{not_1} &=& \swapp \\
\mathsf{not_2} &=& (\swapp \odot \swapp) \odot \swapp 
\end{array}\]
We can write a 2-combinator whose \emph{type} is $\mathsf{not_2}
\isotwo \mathsf{not_1}$:
\[
(\idldl ~\respstwo~ \idisotwo)~\transtwo~\idldl
\]
which not only shows the equivalence of the two implementations of
negation but also shows \emph{how} to transform one to the other.

Fig.~\ref{pi-combinators} lists all the 1-combinators which consist of
base combinators (on the left) and compositions (on the right). Each
line of the base combinators introduces a pair of dual
constants\footnote{where $\swapp$ and $\swapt$ are self-dual.} that
witness the type isomorphism in the middle. This set of isomorphisms
is known to be sound and
complete~\cite{Fiore:2004,fiore-remarks}. Fig.~\ref{pi-combinators2}
lists a few of the 2-combinators that we use in this paper. Each
2-combinator relates two 1-combinators of the same type and witnesses
their equivalence. Both 1-combinators and 2-combinators are invertible
and the 2-combinators behave as expected with respect to inverses of
1-combinators as shown below.

@@

We give an operational semantics for the 1-combinators of $\Pi$ which
represent the conventional layer of programs.  Operationally, the
semantics consists of a pair of mutually recursive evaluators that
take a combinator and a value and propagate the value in the forward
direction~$\triangleright$ or in the backward
direction~$\triangleleft$. We show the complete forward evaluator in
Fig.~\ref{opsem}; the backward evaluator is easy to infer.

@@

\paragraph{Intermezzo.} For an information-theoretic perspective on
the language $\Pi$, we think of a type containing $N$ values as an
abstract system that has~$N$ distinguishable states. According to the
conventional theory of information~\cite{Shannon1948}, the amount of
information contained in each state of a system with $N$
distinguishable states is $\log N$ bits of information. For example,
the type $\mathbb{2}$ can be thought of as an abstract system with
two distinguishable states labeled $\mathsf{true}$ and
$\mathsf{false}$ each containing $\log 2 = 1$ bit. Similarly, the type
$\mathbb{3}$ can be thought of as an abstract system with three
distinguishable states each containing $\log 3$ bits. The logarithmic
map implies that information contained in a composite state is the sum
of the information contained in its constituents. For example, the
type $\mathbb{2} \otimes \mathbb{3}$ can be thought of a composite
system consisting of two independent unrelated subsystems. Each state
of the composite system therefore contains
$\log (2 * 3) = \log 2 + \log 3$ bits which is the sum of the
information contained in each subsystem. Since all the combinators
preserve cardinality, they are also information-preserving.
